---
title: 'Tutorial 1: Getting Started with R and R Studio'
author: "Ana Morales"
date: "`r Sys.Date()`"
fontsize: 12pt
output: 
html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction

This is the first tutorial of the workshop. This is for you to familiarise yourself with R, R Studio and the console.

### 1.1. Installing R and R Studio

The PC in the university cluster have R and R studio already installed, so this task is for installing these software **in your own computer or laptop**. You can skip this section if you already have R and R studio installed.

We will be using R Studio, but R studio can only work if R is already installed, so we need to have both softwares installed. 

- **Installing R**

R is free and can be installed in Windows, Mac, Linux. Go to the [R project website](https://www.r-project.org/). Once there click on **Download R** and select a CRAN Mirror. For the UK you can choose between the 	University of Bristol
Imperial College London, either of them would do the job.  


Now, follow the instructions and you will have R installed in your own computer.  

- **Installing R Studio**

Go to the [R Studio website](https://rstudio.com/products/rstudio/download/), choose the free version and download the file according to your operating system.

Once you have installed both R and R Studio, you don't need to interat at all with R. You can do everything you need from within R Studio.

```{r}

```

## 

You can also embed plots, for example:

```{, echo=FALSE}

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
