---
title: "Exploratory Data Analysis using the Crime Survey for England and Wales"
author: "Ana Morales"
date: "`r Sys.Date()`"
fontsize: 12pt
output: 
html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this practical you will be analysing a sample of the teaching dataset  of the Crime Survey for England and Wales available [here](https://beta.ukdataservice.ac.uk/datacatalogue/studies/study?id=7911) from the [UK Data Service](https://www.ukdataservice.ac.uk/). But you can access it from the folder **Crime_workshop** in the "C:\Work" folder of the PC in the cluster.

## 1. Before we start

#### 1.1. Define a working directory

You can use the directory  on University-owned machines, as it provides a stable drive to work on ("C:\Work"). If you're using your own machine, you can specify any folder you find convenient.

Also note that you need to take your data and results with you and delete what you store on the C-drive after this session (if you are not using your own laptop).  


#### 1.2. Open a new Script

First, open a new script in R studio and save it in your working directory, so you will be able to access this script at a later time if you want to revise or modify a code.  

In R Studio:  

Go to File... New File.... R script
  

#### 1.3. Load the packages

Remember to load the packages. We will be using `tidyverse` and `haven`

```{r}

library(haven)
library(tidyverse)
```





As a first step in a research project we need to start with a purpose:

## Central tendency measures

## Visual exploration

## Association between variables
# Correlation
# Crosstab
# Visual exploration

