---
title: "Exploratory Data Analysis using the Crime Survey for England and Wales"
author: "Ana Morales"
date: "`r Sys.Date()`"
fontsize: 12pt
output: 
  html_document:
    highlighter: null
    #theme: "flatly"
    code_download: TRUE
    toc: TRUE
    toc_float: TRUE
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this practical you will be analysing a sample of the teaching dataset  of the Crime Survey for England and Wales available [here](https://beta.ukdataservice.ac.uk/datacatalogue/studies/study?id=7911) from the [UK Data Service](https://www.ukdataservice.ac.uk/). But you can access it from the folder **Crime_workshop** in the "C:\Work" folder of the PC in the cluster.

# 1. Before we start

### 1.1. Define a working directory

You can use the directory  on University-owned machines, as it provides a stable drive to work on ("C:\Work"). If you're using your own machine, you can specify any folder you find convenient.

Also note that you need to take your data and results with you and delete what you store on the C-drive after this session (if you are not using your own laptop).  


### 1.2. Open a new Script

First, open a new script in R studio and save it in your working directory, so you will be able to access this script at a later time if you want to revise or modify a code.  

In R Studio:  

Go to File... New File.... R script
  

### 1.3. Load the packages

Remember to load the packages. We will be using `tidyverse` and `haven`

```{r, message=FALSE}

library(haven)
library(tidyverse)
```



As a first step in a research project we need to start with a purpose:




# 2. Import the data into R

Import the dataset called hse_data.dta into the console using your desired method (code or point and click). This dataset can only be imported if you have loaded the package `haven`. 

```{r, echo=FALSE}
setwd("C:/Users/Amanda/Desktop/anita/Crime-workshop")

csew <- read_dta("CSEW/UKDA-7911-stata/stata/stata11/csew1314_teaching_eul2.dta")

```

```{r, eval=FALSE}
# Do it yourself

 

```



# 3. Exploratory Data Analysis

## 3.1 Dataset exploration

**Task 1**: Inspect the dataset, use the function View()

```{r, eval=FALSE}

View(csew)

```

You can also use the function `head()` that shows you the first 6 rows of the data set, this function is usefull to have a look at the data within the console, specially when you havelarge datasets.

There is also an antithesis function to `head()` called `tail()`. Can you guess what does that function do?   

(**Hint**: you can run the code in the console and see it by yourself, or a quick search in google may also help.)

```{r, eval=FALSE}
head (csew)
```

Data obtained from the UK Data Service always came with a wealth of resources and user guides that help us to understand the data, so I would always encourgae you to check the documentation associated with the data used.These documentation also includes information about the sample, the questionnaire, the variables availables and how to use them.   

Sadly, this is not always the case with data from different sources, in those cases the initial exploration of the data is probably our main source of information about the data itself.

Now that we got the data in the console, let's start with our data exploration:


**Task 1.1** How many observations does the data have?____________


**Task 1.2** How many variables?______________




Another function used to get a "glimpse" at your data is using the function `glimpse()`.



```{r, eval=FALSE}

glimpse(csew)


```


Note the data type of almost all variables is `<dbl+lbl>`.  These are labelled variables in the haven package. It's the way to identify variables with labels from Stata.



##3.2 Variable exploration 


Let's use another function to have a closer look, this time at the variables. Let's use this with the variable 'sex'
```{r}

attributes(csew$sex)

```


We have the `$class` attribute which says that `sex` is a haven_labelled type. We also have the `$labels` attribute which are the Value Labels of the variable `sex`: 1 Man, 2 Woman

We could have also used `class(hse$sex)` to check the type of the variable.


**Task 2** Inspect other variables with attributes() and class()




# 4. Understanding our data

Now that we have had a look at what variables we have, we need to start looking for other aspects of the exploratory data analysis (EDA) process. These are some of the things to look out for: 

- Checking for outliers (unusual values)
- Looking at the distribution of the variables
- Exploring some relationships and patterns
- Checking for missing cases

## 4.1. Univariate analysis


We can use a series of descriptive statistics and graphs to help us to undertand and make sense of the data.

Univariate exploration is when we look at the characteristics of each variable of interest independently. 

### Univariate Descriptive statistic for categorical (factor) data


Let's see what happens when we ask for a frequency table

```{r}

table(csew$sex)
```

Right now, the variable does not contain any visible label that help us to identify each category of response, although we know the underlying values of each label thanks to the `attribute` function.

This is one of the inconveniences of working with data with labels in R, but it is good for you to know what to do in this case.


Although we can use the data as it is, I prefer to see the values of the variables, so we are going to change the class of some varibles to make them easier to analyse.


We can use the function `as_factor()` from the package **haven** to convert the variable into a factor variable (equivalent to categorical variable). Copy the following code in your console.  

```{r}

csew$sexf<-as_factor(csew$sex) 
table(csew$sexf)
```

Here we created a new variable with the suffix "f" (for factor). 


**Task 3:** Convert the following variables into factor using the example code before.  

- nation: adult respondent nationality
- educat3: respondent education (5 categories)
- ethgrp5a: ethnic group
- bcsvictim: Experience of any crime in the previous 12 months  
    
    

### Missing values


In R missing values are identified as NA.  


Now inspect whether there are missing values in our variables

```{r, echo=FALSE}

csew$nationf<- as_factor(csew$nation)

```

```{r}

#is.na(csew$sexf)

sum(is.na(csew$sexf))
sum(is.na(csew$nationf))

```

The function `is.na` is a logical function which looks at each observation and evaluates whether it is a valid case or a missing case. It will show a series of values `TRUE` for those observation that are missing and `FALSE` for valid cases (is.na = FALSE means that the cases are valid).

It is a good function to explore missingness, but it is not very practical on its own (as you could see in the example) that is why it is often used along other statistics.

I used the statistc `sum` which will add all the `TRUE` values of `is.na` so the result will be the number of NA (missing cases) for a particular variable.

The data shows that there are no missing values, but we can infer that the values 8 and 9 can be classifyed as missing cases, so we recode them as missing

```{r}
csew$nationf<- recode(csew$nationf, "Refused" = "NA", "Don't know" = "NA")

table(csew$nationf)

```

We have succesfully recode those values as missing, now we need to tell R Studio to treat those values as missing too

```{r}
csew$nationf<- na_if(csew$nationf, "NA")

csew$nationf<- droplevels(csew$nationf) #with this code we got rid of the category for NA
table(csew$nationf)

```

**Task 4:** Check for missing values in the following variables and deal with them accordingly:

- educat3: respondent education (5 categories)
- ethgrp5a: ethnic group
- bcsvictim: Experience of any crime in the previous 12 months



### Univariate Descriptive statistic for continuous (numeric) data

We can get summaries of the main statistics using the base function `summary()`. This function gives us the minimum and maximun values, the mean, median and the quartiles.   

It is really useful to check for extreme values. Let's have a go

```{r}
#This uses basic functions from R
summary(csew$age)
```

`summary()` does not gives any measure of the spread of the, we can use other functions available in base R for that `sd()` for standard deviation.

```{r}
# Standard deviation
sd(csew$age)

```

so far we have used two different functions of base R for exploratory data analysis. We can obtain the same statistics and more using a combination of different function from the package `tidyverse` taking advantage of the `%>%` (pipe) operator from the package **magrittr** which is fully implemented in tidyverse, so there is no need to install anything else.


I'm a firmer believer (although I made be wrong) that the best way to learn is putting into practice, so this is a perfect opportunity to put `%>%` into practice. (You can check your notes of tutorial 1 for more details on how to use them).

Let's start with the `summarise` function from the package `dplyr` (also part of tidyverse) This function allows us to create summary statistic of our variables in one single code


```{r}

#using function from the package dplyr (tidyverse)
csew %>% 
  summarise(mean_age= mean(age)) # we create the variable mean_age


```

That was good but not very impressive, is it? We can add more function to that code and ask for more:  

```{r}

csew %>% 
  summarise(mean_age= mean(age), 
            sd_age= sd(age),
            median_age = median(age),
            min_age = min(age),
            max_age = max(age)) 

```

One of the advantages of using `%>%` is that we can add more functions to our code as a set of instructions. In the exaple below, I added the function `filter()` of the package `dplyr` to exclude all missing cases from the analysis.  


The function `is.na()` (is missing?) will select all missing cases, but when used along the `!` it means totally the opposite, so in this code this stand for **filter all the cases of variable age that *are not* missing**


```{r}
csew %>%
  filter(!is.na(age)) %>%
  summarise(mean_age= mean(age), 
            sd_age= sd(age),
            median_age = median(age),
            min_age = min(age),
            max_age = max(age)) 

```

We can even add more variables! but for that we might need another function and the chance to use more pipes %>%.

- [select](https://dplyr.tidyverse.org/reference/select.html): This function allows to select specific variable from the datasets.  
- [summarise_all](https://dplyr.tidyverse.org/reference/summarise_all.html): allow to apply the same transformation to multiple variables
- `list`: used to make a list of functions to apply, in this example, mean and sd


```{r}
csew %>% 
  select(age, antisocx) %>%
  summarise_all(list(mean= mean, sd= sd, median = median), na.rm = TRUE)

```
What can we said about age?  

The distribution of the variable age seems reasonable, the mean and the median are very similar. 

The standard deviation shows how spread the data are. A standard deviation of 57.68 is quite high, which might indicate a lot of variation in the distribution of age. This is clearly hinting that there is something odd about the variable.

We also saw that the maximum value for age is **999**, which is clearly a number out of range; Nobody can live for 999 years! This is probably the reason why the standard deviation is so high, **we clearly need further exploration of this unusual value**

In certain cases the maximun values can be less obvious, so it would be difficult to determine whether there are outliers or data errors, that is why we need to complement this intial checkings with visual exploration.

**Diagnostic, critical points regarding age**

- high standard deviation
- Implauible values (errror?, missings?, outliers?)

**Actions**

- both problems might be related
- check 999 values and deal with them accordingle
- check if SD changes after 999 values are took into account

```{r}
class(csew$age)
attributes(csew$age)

```

using the fucntion `attributes()` we now know  

- age is formated as "%8.0g" which is a way of storing numeric data in Stata. 
- age is a "haven_llabelled" variables, which means there are certaing values with labels
- The values and labels 998 and 999 are for "refused" and "don't know" respectivelly.

998 and 999 can be classifyed as missing cases (unless you are interested in knowing the patterns of non-response). So we can easily solve that problem.


```{r}
csew$age[csew$age==998 | csew$age==999 ]<-NA 

```


**Task 5:** Get summary statistics for the variables **confx**, **fair** and **effectx**. Make note of any pattern or unusual value. A description of the variables can be found in this [user guide](http://doc.ukdataservice.ac.uk/doc/7911/mrdoc/pdf/7911_csew_2013-14_teaching_dataset_user_guide.pdf)


**Task 5.1** Write down a short description of the main statistics you've obtained



**Task 5.2** Is there any unusual value in the variables you have analysed?




## 4.2. Visual exploration

We can use a combination of bar plots, histograms, boxplots and scatterplots as the basis for our data exploration

**bar plots of categorical data**
```{r}
 
ggplot(data = csew) +
  geom_bar(mapping = aes(x = sexf))
```

**histogram for continuous data**

```{r}
ggplot(data = csew) +
  geom_histogram(mapping = aes(x=age), binwidth = 0.5, na.rm = TRUE)
```

Both bar chart and histograms can be used to detect outliers and unusual patterns of the data. 

**Task 6:** Visually explore the variables that we have used in the previous task. Make note of any unusual patter that you can find.



## 4.3. Association between variables

More insights can be gained when lookig at relationships between variables. This can be done in two main ways: using numerical representation of this association and visual representations.

### Associations using "crosstab" or "contingency" tables

For categorical data we can use contingency tables (also called crosstabs)

Here we will be using a new package called `janitor`.


```{r, message=FALSE}
#install.packages("janitor") 
library(janitor)

```

```{r}
csew %>%
  tabyl(sexf, nationf, show_na = FALSE) %>% #excludes NA values from the table
  adorn_percentages("row") %>% # takes row percentage
  adorn_pct_formatting(digits = 2) %>% # limit % to 2 digits
  adorn_ns() #add the cell count


```



**Task 7:** Explore the relationship between victim of crime and sex, ethnic group, nationality and educational level. Make notes of your findings




### Looking at association through visualisations

Now we need to focus on our variable of interest (outcome usually represented as the 'y' variable) and its association with other explanatory variables (x variables).

Here we will be looking at the associations between confidence in the effectivness of Criminal Justice System (effectx) and other individual characteristics of the respondents.

**Two continous variables**


-Scatter plots: For looking at the relationship between two continous variables. This is usuful to assess co-variation

```{r}

ggplot(data = csew, mapping = aes(x = age, y = effectx)) + 
  geom_point(na.rm = TRUE)


```

We received a warning regarding a labelled variable, further exploration reveals that the variable age is numeric but contains two labels ("haven_labelled") for "refused" and "don't know".

The `as.numeric()` function will create a copy of the variable age of class "numeric". The function will get rid of the labels.

```{r}

csew$age2<-as.numeric(csew$age)

```

You can check that we now got rid of the warning.

**Categorical and continous variables**

- boxplot: ideal for detecting outliers. Here we obtain boxplots by groups

```{r}
ggplot(data = csew, mapping = aes(x = sexf, y = effectx)) +
  geom_boxplot(na.rm = TRUE)
```

We can add colours to the ouliers

```{r}
ggplot(data = csew, mapping = aes(x = sexf, y = effectx)) +
  geom_boxplot(outlier.colour = "red", na.rm = TRUE)
```


**Task 8:** Explore the associations between effectivenesss in the Criminal Justice System and nationality, ethnic group, victim of crime




**Task 9:** After all the bivariate exploration that you have done. Have you found any difference in the perception of the effectiveness in the Criminal Justice System according to:

- sex____________________________________________________________
_________________________________________________________________
- ethnic group: _________________________________________________
_________________________________________________________________
- Victims: ______________________________________________________
_________________________________________________________________


**Task 10:** Can you think in any hypothesis that can help us to understand the differences in the perception of how effective the Criminal Justice System is? 




